Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-26 trunk/paper/drinking.tex

\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
  A. Harris$^3$ and ... and Gabriella Rustici$^2$ and Georgios
  V. Gkoutos$^{1}$}

@@

While the PATO framework is now successfully being applied to
semantically integrate phenotypes across species, the diversity and
complexity of phenotypes in which biological processes and functions
are impaired continues to limit the interoperability between phenotype
ontologies. Biological processes and functions are linked to the
physiology which the components of biological systems exhibit and

open: process phenotypes very complex...  
physiology...
measurements and inferences...

\subsection{Problem statement 1}
Heart rate increased: frequency of CMC increased fluid flow increased
volume per contraction increased ...

@@

\subsection{Problem statement 2}
increased apoptosis

\subsection{Observations}
observation 1: there is a fundamental distinction between attribution
of qualities for a fixed period (e.g., 24h) and single {\em act}s of
drinking.

>>>>

\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
  A. Harris$^3$ and Heinrich Herre(?) and Gabriella Rustici$^2$ and
  Georgios V. Gkoutos$^{1}$}

@@

While the PATO framework is now successfully being applied to
semantically integrate phenotypes across species, the diversity and
complexity of phenotypes in which biological processes and functions
are impaired continues to limit the interoperability between phenotype
ontologies. Major challenges for representing and integrating
process-based phenotypes include establishing the link to the
components of biological systems that have the capabilities to exhibit
such a behaviour, and that attributes of processes are often measured
{\em indirectly} and inferred from other attributes.

To illustrate these challenges, consider physiological processes of
the heart. One of the heart's functions is {\em Heart beating}, i.e.,
a capability that is realized through processes of the type {\em Heart
  beating}. {\em Blood} is a participant of {\em Heart beating}
processes.  An abnormal phenotype of an organism could be that the
{\em rate of heart beating} is increased with the intended meaning
that the number of {\em Heart beating} processes in a given time
interval is higher than normal. An attribute such as {\em increased
  rate of heart beating}

Heart rate increased: frequency of CMC increased fluid flow increased
volume per contraction increased ...

@@

%%%%%%%%%%%%%%%%%%
observation 1: there is a fundamental distinction between attribution
of qualities for a fixed period (e.g., 24h) and single {\em act}s of
drinking.
